SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-01-15 paper/paper.tex
Paper updates.

For example, consider a simple calculator language containing constants,
addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
Consider a program that takes an \ttp{Exp} as an argument that throws an error
if an evaluation of an \ttp{Exp} value contains a division-by-zero.  Testing for
the occurrence of the error, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (C 3) (Add
(C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
In the counterexample, the culprit is a deeply-buried sub-value,
%
\begin{code}
Div (C 3) (Add (C (-5)) (C 5))
\end{code}
%
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  (The size of the
value is with respect to the \ttp{size} method in the \ttp{SubTypes} class.)
Finally, \ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) =
  1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

The generalization step is an important under-appreciated step in the debugging
process.  For counterexamples that are large, deeply-nested values, the
generalization step can be difficult.  One way we might help the programmer is
to provide her not with a single counterexample but with a quantified formula
that characterizes a set of counterexamples.  Doing so has two benefits.  First,
it already generalizes a counterexample, perhaps helping the programmer's
insight in the generalization step.  Second, it reduces the noise in favor of
the signal by abstracting away portions of large counterexample that are
irrelevant to why it violates the property.

The characterization of counterexamples that most helps the programmer seems to
should strike a middle ground.  At the one extreme, a single counterexample is
can be thought of as a closed formula containing only literals that
under-approximates the set of counterexamples.  At the other extreme, the
property itself being tested is a formula that over-approximates the failing
inputs.  We describe two kinds of formula that fall between these two extremes.

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so
significantly.  We call the kind of formula just shown \emph{universal sub-value
  generalization} and is implemented in SmartCheck.

An extrapolation algorithm performs universal sub-value generalization.  The
basic idea is as follows: for a counterexample \ttp{cex} and a property
\ttp{prop}, a breadth-first search over the sub-values of the \ttp{cex} is
performed.  For each sub-value, the algorithm generates new sub-values, replaces
them in \ttp{cex} to create a list of new potential counterexamples.  If enough
new counterexamples non-trivially fails \ttp{prop} (i.e., does not just fail the
precondition), and non satisfy \ttp{prop}, then we extrapolate, claiming that
for \emph{any} new value replacing that sub-value in \ttp{cex}, the property
will fail.

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard in which a call to \ttp{subTree}
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\ttp{subTree cex idx0 idx1} is true if in \ttp{cex}, the value at index
\ttp{idx0} is a child of index \ttp{idx1}, in a tree representation of the
data.  This prevents the algorithm from trying to generalize sub-values that are
abstracted away already since their parents have been generalized.

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if enough
new values pass the property's precondition and fail the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is the roughly the dual of \ttp{pass} in the reduction algorithm: \ttp{fail prop
  cex} returns \ttp{Just True} if \ttp{cex} passes \ttp{prop}'s precondition but
fails the property; \ttp{Just False} if \ttp{cex} non-trivially satisfies
\ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails \ttp{prop}'s precondition.  The
constant \ttp{scMinExtrap} is the minimum number of \ttp{Just False} results
required from \ttp{fail} to extrapolate from failed tests to a universal claim.

@@

Sum types denote choice in a data type.  Sometimes, a property over a sum type
fails because there is a bug for some of the variants but not others.  For
example, recall the calculator language from Section~\ref{sec:subval}.  The
property \ttp{prop\_div} fails only for values that contain a variant tagged
with the \ttp{Div} constructor.  For example, consider the following counterexample:
%
\begin{code}
Div (Add (C 20) (C 30))
    (Add (C 89) (C -89))
\end{code}
%
\noindent
The counterexample fails because the divsor evaluates to zero.  Using universal
sub-value generalization described above, the counterexample is generalized as
follows:
%
\begin{code}
forall x . Div x (Add (C 89) (C -89))
\end{code}
%
\noindent
So we know there is something particular about the \ttp{Add (C 89) (C -89)},
because there are some values that we can replace this sub-value with in the
original counterexample causing it to satisfy the property.  But we might wonder
if there is something special about variants tagged by the \ttp{Add}
constructor, or might we finding failing sub-values with the other variants
tagged by the constructors \ttp{C} and \ttp{Div}?  If so, then we know there is
not some bug being triggered by \ttp{Add} variants specifically, so it is more
likely that the bug involves how the second field of the \ttp{Div} variant is
evaluated.

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following all characterize counterexamples:
%
\begin{code}
forall x . Div x (Add (C 89) (C -89))
forall x . Div x (C 0)
forall x . Div x (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the second field of the outermost \ttp{Div}
constructor.  We therefore return a formula like the following:
%
\begin{code}
forall values x .
  forall constructors C\(\sb{0}\) .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (C\(\sb{0} \stackrel{\rightarrow}{y}\))
\end{code}
%

>>>>

For example, consider a simple calculator language containing constants,
addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
Consider a program that takes an \ttp{Exp} as an argument that throws an error
if an evaluation of an \ttp{Exp} value contains a division-by-zero.  Testing for
the occurrence of the error, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (Add (C 7)
  (C 3)) (Add (C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
In the counterexample, the culprit is a buried sub-value,
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  (The size of the
value is with respect to the \ttp{size} method in the \ttp{SubTypes} class.)
Finally, \ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length, and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) =
  1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

The generalization step is an important under-appreciated step in the debugging
process.  For counterexamples that are large, deeply-nested values, the
generalization step can be difficult.  One way we might help the programmer is
to provide her not with a single counterexample but with a quantified formula
that characterizes a set of counterexamples.  In particular, a formula reduces
the noise in favor of the signal by abstracting away portions of large
counterexample that are irrelevant to why it violates the property.

The characterization of counterexamples that most helps the programmer should
strike a middle ground.  A single counterexample is too specific.  On the other
hand, the property itself is a formula that over-approximates the failing
inputs.  We describe two kinds of formula that fall between these two extremes.

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
An example best explains the first kind of quantified formulas we generate.
Recall the calculator language from Section~\ref{sec:subval}.  Recall the
culprit buried within a counterexample, causing it to result in a
division-by-zero error:
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Notice that any value substituted in the dividend results in an error.  We can
quantify the error then as follows:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Not only does the generalization assist the programmer's insight, but it reduces
the sheer size of the counterexample, and it can do so significantly.  We call
the kind of formula just shown \emph{universal sub-value generalization} and is
implemented in SmartCheck.

An extrapolation algorithm performs universal sub-value generalization.  The
basic idea is as follows: for a counterexample \ttp{cex} and a property
\ttp{prop}, a breadth-first search over the sub-values of the \ttp{cex} is
performed.  For each sub-value, the algorithm generates new sub-values, replaces
them in \ttp{cex} to create a list of new potential counterexamples.  If enough
new counterexamples non-trivially fail \ttp{prop} (i.e., fail the property's
consequent but satisfy its antecendent), then we extrapolate, claiming that for
\emph{any} new value replacing that sub-value in \ttp{cex}, the property will
fail.

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard in which a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
is true if in \ttp{cex}, the value at index \ttp{idx0}, is a child of index
\ttp{idx1}, in a tree representation of the data.  This prevents the algorithm
from trying to generalize sub-values that are abstracted away already since
their parents have been generalized.

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if enough
new values pass the property's precondition and fail the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is the roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.  The constant \ttp{scMinExtrap} is the minimum number
of \ttp{Just False} results required from \ttp{fail} to extrapolate from failed
tests to a universal claim.

@@

Sum types denote choice in a data type.  Sometimes, a property over a sum type
fails because there is a bug for some of the variants but not others.  For
example, recall the calculator language from Section~\ref{sec:subval}.  The
property \ttp{prop\_div} fails only for values that contain a variant tagged
with the \ttp{Div} constructor.  Recall again the generalized counterexample
from Section~\ref{sec:generalization}:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Because the divisor does not generalize, we know there is something special
about it that causes failure.  But we might wonder if there is something special
about variants tagged by the \ttp{Add} constructor, or might we finding failing
sub-values with the other variants.

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following all characterize counterexamples:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
forall x . Div x (C 0)
forall x . Div x (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore return a formula like the
following:
%
\begin{code}
forall values x .
  forall constructors C\(\sb{0}\) .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (C\(\sb{0} \stackrel{\rightarrow}{y}\))
\end{code}
%
